Unbalanced Mallows Models for Optimizing Expensive Black-Box Permutation Problems

### 2020-10-27 paper/main.tex
fix intro
fix background

In this work, we propose and analyze Unbalanced Mallows Model (UMM), a
population-based probabilistic algorithm specifically designed for optimizing
black-box problems over a permutation landscape under a limited budget. There
exists a wide body of research on probabilistic models for permutations including  Plackett-Luce and Mallows
algorithms~\citep{???},\MANUEL{what can we cite from that body of research?: Ekhine will fix this.}
with the most prominent examples being estimation of distribution algorithms
(EDAs)~\cite{Ceberio2013}.  ACO may also be considered a type of EDA, since
it builds a probability distribution model from which solutions are sampled
with a bias towards the best solutions evaluated so far. The proposed UMM
drastically differs from EDAs since EDAs assume that function evaluations are
cheap and they can do many of them. On the contrary, UMM aims at learning as
much as possible with a limited number of function evaluations. Moreover,
unlike EDAs, UMM does not rely on a fixed-size population of solutions that
evolves over ``generations''. Instead, taking inspiration from Bayesian
optimizers, UMM considers a sample of permutations during the whole execution
that is increased by one new permutation at each iteration. Finally,
the balance between intensification and exploration is dynamically adapted by
controlling the variance of the probabilistic Mallows model.  \manuel{}{
  \begin{itemize}
  \item Who else has used Mallows model for black-box optimization? Ekhine: Va a mirarlo.
  \item Who else has used unbalanced borda for optimization? Ekhine va a añadir 2 frases.
  \end{itemize}
}

@@

\subsection{Unbalanced Borda}\label{sec:uborda}
We introduce now the core for our probabilistic algorithm, the uBorda algorithm. Following the convention of the community, the ranking returned by the uBorda algorithm will be denoted uBorda ranking. The uBorda algorithm is a recent generalization of Borda~\cite{???}.\MANUEL{X This paper is missing \ekhine{}{to be pocessed by arxiv}} In uBorda, we consider that each permutation in the sample $\sigma\in S$ has an associated weight $w(\sigma)$ where $w\colon S_n \to \mathbb{R^{+}}$. Then, the score of \manuel{item}{position} $i$ is defined as follows:
%
\begin{equation}\label{eq:uborda_score}
B(i) =  \sum_{t\in S}  w(\sigma_t) \sigma_t(i) \enspace,
\end{equation}
%
and, the same as Borda, orders the \ekhine{items}{positions}\MANUEL{X why items? we talk about positions above. We could talk about decision variables if that is clearer\ekhine{}{id rather use the simplest possible notation. In statistical literature you ll see items/rankings. }} increasingly by their score.

Borda and uBorda are equivalent when $w(\sigma)$ is constant for every $\sigma$. However, in other case uBorda ranking will be closer to the rankings in $S$ with smaller value of $w(\sigma)$. Roughly speaking, it is equivalent to replicating in the sample the $\sigma$ with low values of $w(\sigma$).
Clearly, the choice of  $w(\sigma)$ depends on the relevant property of a ranking in a particular domain. For example, uBorda has been shown to have nice theoretical properties for the online estimation of the parameters of a sample with concept drift~\cite{}.\MANUEL{X where? Do we miss a citation here?\ekhine{}{to be processed, arxiv} Maybe you can create a bibentry with the details that you know already.}

In this paper, we would like to replicate each permutation $\sigma$ a number of times that is inversely proportional to $f(\sigma)$. Therefore, we will use $w(\sigma)=\rho^{f(\sigma)}$ for a $\rho\in(0,1)$. Interestingly, uBorda is computed in polynomial time.

>>>>

In this work, we propose and analyze Unbalanced Mallows Model (UMM), a
population-based probabilistic algorithm specifically designed for optimizing
black-box problems over a permutation landscape under a limited budget. There
exists a wide body of research on probabilistic models for permutations including  Plackett-Luce and Mallows
algorithms~\citep{???},\MANUEL{what can we cite from that body of research?: Ekhine will fix this.}
with the most prominent examples being estimation of distribution algorithms
(EDAs)~\cite{Ceberio2013}.  ACO may also be considered a type of EDA, since
it builds a probability distribution model from which solutions are sampled
with a bias towards the best solutions evaluated so far. The proposed UMM
drastically differs from EDAs since EDAs assume that function evaluations are
cheap and they can do many of them. On the contrary, UMM aims at learning as
much as possible with a limited number of function evaluations. Moreover,
unlike EDAs, UMM does not rely on a fixed-size population of solutions that
evolves over ``generations''. Instead, taking inspiration from Bayesian
optimizers, UMM considers a sample of permutations during the whole execution
that is increased by one new permutation at each iteration. Finally,
the balance between intensification and exploration is dynamically adapted by
controlling the variance of the probabilistic Mallows model.  \manuel{}{
  \begin{itemize}
  \item Who else has used Mallows model for black-box optimization? Ekhine: Va a mirarlo.
  \item Who else has used unbalanced borda for optimization? Ekhine va a añadir 2 frases.
  \end{itemize}
}
This paper structured as follows \MANUEL{Mío}

@@

\subsection{Unbalanced Borda}\label{sec:uborda}
We introduce now the core for our probabilistic algorithm, the uBorda algorithm. Following the convention of the community, the ranking returned by the uBorda algorithm will be denoted uBorda ranking. The uBorda algorithm is a recent generalization of Borda~\cite{???}.\MANUEL{X This paper is missing \ekhine{}{to be pocessed by arxiv}} In uBorda, we consider that each permutation in the sample $\sigma\in S$ has an associated weight $w(\sigma)$ where $w\colon S_n \to \mathbb{R^{+}}$. Then, the score of item $i \in [n]$ is defined as follows:
%
\begin{equation}\label{eq:uborda_score}
B(i) =  \sum_{t\in S}  w(\sigma_t) \sigma_t(i) \enspace,
\end{equation}
%
and, the same as Borda, orders the items $i \in [n]$ increasingly by their score.

Borda and uBorda are equivalent when $w(\sigma)$ is constant for every $\sigma$. However, in other case uBorda ranking will be closer to the rankings in $S$ with smaller value of $w(\sigma)$. Roughly speaking, it is equivalent to replicating in the sample the $\sigma$ with low values of $w(\sigma$).
Clearly, the choice of  $w(\sigma)$ depends on the relevant property of a ranking in a particular domain. For example, uBorda has been shown to have nice theoretical properties for the online estimation of the parameters of a sample with concept drift~\cite{}.\MANUEL{X where? Do we miss a citation here?\ekhine{}{to be processed, arxiv} Maybe you can create a bibentry with the details that you know already.} 
%
% In this paper, we would like to replicate each permutation $\sigma$ a number of times that is inversely proportional to $f(\sigma)$. Therefore, we will use $w(\sigma)=\rho^{f(\sigma)}$ for a $\rho\in(0,1)$.
Interestingly, uBorda is computed in polynomial time.
